A general model for the evolution of nuptial gift-giving

### 2024-12-20 index.tex
Update last comments based on round of revisions and Evol Lett comments

We develop a general framework for investigating the evolution of
nuptial gift-giving and choosiness using a time-in, time-out modelling
approach and an individual-based model
(\protect\hyperlink{ref-Clutton-Brock1992}{Clutton-Brock \& Parker,
1992}). Specifically, we derive conditions under which selection will
favour male search for nuptial gifts and female rejection of gift-less
males. We show that selection for searching and choosiness depends on
whether a threshold fitness value of the nuptial gift is exceeded. Our
model demonstrates the importance of nuptial gift cost, sex ratio, and
mate encounter rate in determining the threshold above which selection
will favour the evolution of nuptial gift-giving. Importantly, we show
that the threshold value differs for male searching and female
choosiness. We complement the predictions of our analytical model by
formulating an individual-based model, which further supports the main
theoretical results. We apply our model to an example system with
nuptial gifts, the nursery web spider \emph{Pisaura mirabilis}, where we
use experimental data to estimate a key model parameter. Our results
provide a general framework for understanding why nuptial gift-giving
evolves in some systems and not in others, how the evolution of nuptial
gift-giving can give rise to sexual conflict, and it provides insight
into the evolution of worthless and deceitful nuptial gifts, which occur
in several different taxonomic (\protect\hyperlink{ref-LeBas2005}{LeBas
\& Hockham, 2005}; \protect\hyperlink{ref-Ghislandi2014}{Ghislandi
\emph{et al.}, 2014}).

@@

\begin{figure}
\includegraphics[width=1\linewidth]{inst/img/conceptual_figure} \caption{Conceptual figure inspired by Kokko and Ots (2006) illustrating how the modelling framework maps onto an example of a system wherein nuptial gifts are used, here \textit{Pisaura mirabilis}. Males have a probability of obtaining a nuptial gift while in time-out, which will affect their probability of mating while in time-in. They return to the mating pool (time-in) at a rate determined by the time spent searching for a nuptial gifts ($T_{\mathrm{m}}$) and leave the mating pool (i.e. enter time-out) following the female encounter rate, which is dependent on the ratio of males to females ($\beta$) and the encounter rate ($R$). The choosy sex (females) enter the mating pool at a rate depending on the time spent processing offspring ($T_{\mathrm{f}}$) and leave the mating pool (i.e. enter time-out) at a rate that is dependent on $\beta$ and $R$. Males and females undergo sex-specific mortality $\mu$ during time-in and time-out. Image left to right: (1) male \textit{P. mirabilis}. (2) male \textit{P. mirabilis} presenting nuptial gift (white) to female. (3) Female \textit{P. mirabilis} protecting offspring. Photos: Alamy.}\label{fig:unnamed-chunk-2}
\end{figure}

@@

The probability that a male succeeds in securing a nuptial gift (\(G\))
after a duration of \(T_{\mathrm{m}}\) is defined by,

\[P(G) = 1 - e^{-\frac{1}{\alpha}T_{m}}.
\tag{1}
\]

In Eq. 1, \(\alpha\) defines expected search time before encountering a
nuptial gift. Thus, the probability of finding a nuptial gift is higher
the more time is spent searching. During time-in, individuals encounter
conspecifics at a rate of \(R\). A focal individual will therefore
encounter conspecifics of the opposite sex at a rate of \(R/2\) if the
ratio of males to females in the mating pool (\(\beta\)) is equal. More
generally, males will be encountered at a rate of \(R\beta/(\beta+1)\),
and females will be encountered at a rate of \(R/(\beta+1)\). An example
of how the structure of the time-in time-out model applies to a system
with nuptial gift-giving is given in Figure 1. We assume that mating
with a nuptial gift increases expected offspring production by an
increment of \(\gamma\). Mortality occurs for females and males in
(\(\mu_{\mathrm{i,f}}\), \(\mu_{\mathrm{i,m}}\)) and out
(\(\mu_{\mathrm{o,f}}\), \(\mu_{\mathrm{o,m}}\)) of the mating pool.
Following Kokko \& Ots (\protect\hyperlink{ref-Kokko2006}{2006}), we
assume \(m_{\mathrm{i,f}} = m_{\mathrm{o,f}} = 1\) and
\(m_{\mathrm{i,m}} = m_{\mathrm{o,m}} = 1\).

@@

During time-out, males can search for a nuptial gift. Males can adopt
one of two strategies; either search or do not search for a nuptial
gift. Males with the searching strategy continue to search until they
find a nuptial gift, while males that do not search immediately re-enter
the mating pool. Consequently, time searching for a nuptial gift will
come at the cost of mating opportunities but might increase offspring
fitness. We therefore need to model the expected length of time spent
outside of the mating pool for males that search for nuptial gifts
(\(E[T_{\mathrm{m}}]\)), which equals \(\alpha\). We can integrate
search time \(t\) over the rate at which nuptial gifts are encountered
(\(\exp(-1/\alpha)\)) to show \(E[T_{\mathrm{m}}] = \alpha\),

@@

When this inequality holds, males are favoured to search until they find
a nuptial gift, which would result in an average search time of
\(\alpha\). If the male trait is instead continuous (i.e., males search
for time period \(T_{\mathrm{m}}\)), it can be shown that the same
threshold can be reached by evaluating the partial derivative of the
male fitness function (Supporting Information S1). Hence, the thresholds
are consistent under different assumptions concerning male searching
strategy. Selection will cause males to search for nuptial gifts if the
fitness increase to offspring exceeds the product of search time and
female encounter rate.

@@

We then evaluate the conditions under which
\(W_{\mathrm{F_{G}}} > W_{\mathrm{F_{G,L}}}\). We can isolate \(\gamma\)
to determine how much offspring fitness must be increased to make
choosiness beneficial (\(\gamma_{\mathrm{f}}\)),

@@

In Eq. 6, \(P_{IM}(t)\) and \(P_{IF}(t)\) are the probabilities of
finding a male and female in time-in, respectively. There is no closed
form solution to the operational sex ratio, so we used recursion to
calculate \(\beta\) values for a given \(T_{\mathrm{f}}\),
\(T_{\mathrm{m}}\), and \(R\) (see Supporting Information S2),

@@

\[P(G^{i}) = 1 - e^{-\frac{1}{\alpha}T^{i}_{\mathrm{m}}}.\]

@@

We ran simulations in which male search time and female choosiness
evolved from an ancestral state of no searching and no choosiness. In
all simulations, \(N\) was initialised at 1000 and \(K = 1000\).
Simulations ran for \(t_{max} = 10000\) time steps. We set
\(T_{\mathrm{f}} = 2\), \(\psi = 3\), and \(\lambda = 1\) for all
simulations, and we simulated across a range of
\(\alpha = \{0.1, 0.2, ..., 1.9, 2.0\}\) and
\(\gamma = \{0, 0.1, ..., 1.9, 2.0\}\) parameter values for 3000
replicates. Summary statistics for mean trait values, population size,
sex ratios, proportion of females and males in and out of the mating
pool, and mean number of encounters per female and male within the
mating pool were all calculated in the last time step. The C code used
for simulating these IBMs also allows for the reporting of summary
statistics in each time step. Additionally, it can simulate explicit
space and individual movement through the landscape. A neutral evolving
trait was also modelled to ensure that the code functioned as intended,
and processes were compartmentalised into individual functions to
facilitate code testing. All code is publicly available on GitHub
(\url{https://github.com/bradduthie/nuptial_gift_evolution}).

@@

\begin{figure}
\centering
\includegraphics{index_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{Fitness thresholds above which males increase their fitness by
searching for nuptial gifts (blue lines; Eq. 2) and females increase
their fitness by rejecting males that do not offer gifts (red lines; Eq.
3). Parameter space includes areas in which males do not search for
nuptial gifts and females are not choosy (A), males search but females
are not choosy (B), females would be choosy but males do not search (C),
and males search and females are choosy (D). Arrows in panel a indicate
the effect of increasing encounter rate (\(R\)), female time-out
(\(T_{\mathrm{f}}\)), and male search time (\(T_{\mathrm{m}}\)). As an
example, trajectories for \(T_{\mathrm{f}} = 2\), and
\(T_{\mathrm{m}} = \alpha\) are shown for values of \(R=2\) (panel a)
and \(R = 1\) (panel b). Females are assumed to be the choosy sex, which
is maintained as long as \(\alpha < T_{\mathrm{f}}\).}
\end{figure}

For males, if nuptial gifts are not abundant and thus require a long
time to find (i.e., high \(\alpha\)), or if males encounter many females
per unit time (i.e., high \(R / (1+\beta)\)), then the nuptial gift must
result in a high fitness increment for selection to favour searching.
For females, if \(\gamma\) is sufficiently high, fitness is increased by
rejecting males without gifts and mating only with males that provide
nuptial gifts. As offspring processing time (\(T_{\mathrm{f}}\)), mate
encounter rate (\(R\beta / (\beta + 1)\)), or the probability of a male
finding a nuptial gift (\(1 - \exp(-T_{\mathrm{m}}/\alpha)\) ) decrease,
the threshold value of fitness above which selection will favour
choosiness (\(\gamma_{\mathrm{f}}\)) increases. This can be understood
intuitively by realising that rejecting a prospective male represents an
opportunity cost for the female.

@@

We show that the threshold nuptial gift value at which females are
favoured to express choosiness for nuptial gifts is rarely equivalent to
the threshold value at which males are favoured to search for nuptial
gifts, potentially leading to sexual conflict
(\protect\hyperlink{ref-Arnqvist2005a}{Arnqvist \& Rowe, 2005};
\protect\hyperlink{ref-Oliveira2008}{Oliveira \emph{et al.}, 2008}).
Here, we are defining sexual conflict as occurring when interactions
between sexes result in situation where both sexes cannot achieve an
optimal outcome simultaneously
(\protect\hyperlink{ref-Parker2006}{Parker, 2006}). As an example,
sizable areas of parameter space exists wherein the female optimum would
be to exhibit preference for (and receive) nuptial gifts, while the male
optimum is to not search for (and give) nuptial gifts (see Figure 2a,
Zone C). This will lead to mate encounters in which gift-less males will
benefit from mating, but females will not (i.e., evolutionary interests
do not always overlap with respect to mating strategy between the
sexes). In many systems, ecological variables such as search time
required to find a nuptial gift will likely depend on prey abundance,
which can vary substantially with time in some species with nuptial
gift-giving (\protect\hyperlink{ref-Ghislandi2018}{Ghislandi \emph{et
al.}, 2018}). Since several ecological variables likely affect the value
of these thresholds, our results can be seen as providing some
formalised description of why nuptial gift-giving only occurs in some
but not all systems.

At first, the analytical model seems to suggests that nuptial gifts must
cause a very high fitness increase (approximately 25\%) before male
search and female choosiness is favoured by selection (Figure 2).
Similarly the IBM model seems to suggest that a fitness benefit of
approximately 50\% is required (see Figure 3). However, it is important
to recognise that these thresholds depend on multiple parameters. For
example, if female processing time (\(T_{\mathrm{f}}\)) is high, the
female threshold for choosiness with respect to \(\gamma\) drops such
that male search and female choosiness are favoured at lower \(\gamma\).
If \(T_{\mathrm{f}}\) is sufficiently high, then an initially rare
gift-giving trait might be favoured by selection even if the fitness
benefit of a nuptial gift is low. The effect that nuptial gifts have on
fitness might vary across species, or even populations. Effects on
female fecundity have been estimated in crickets, fireflies,
butterflies, and spiders, but these estimates vary considerably between
species suggesting a large positive effect to no effect at all
(\protect\hyperlink{ref-Bergstrom2002}{Bergström \& Wiklund, 2002};
\protect\hyperlink{ref-Rooney2002}{Rooney \& Lewis, 2002};
\protect\hyperlink{ref-Maxwell2018}{Maxwell \& Prokop, 2018};
\protect\hyperlink{ref-Gao2019}{Gao \emph{et al.}, 2019}).

@@

@@

@@

@@

@@

@@

\hypertarget{s1-alternative-derivation-of-male-fitness-threshold}{%
\subsection{S1: Alternative derivation of male fitness
threshold}\label{s1-alternative-derivation-of-male-fitness-threshold}}

In the main text, we assumed that males made the decision to search or
not search for a nuptial gift. The expected length of time for which
searching males are expected to remain outside of the mating pool is
\(E[T_{\mathrm{m}}] = \alpha\) (see Model). Alternatively, we can assume
that males search for a period of \(T_{\mathrm{m}}\) and spend this full
duration of \(T_{\mathrm{m}}\) in the time-out phase, even if they
succeed in finding a nuptial gift (as we assumed in our IBM). The
probability that a male obtains a nuptial gift during this time is
modelled in Eq. 1,

\[P(G) = 1 - e^{-\frac{1}{\alpha}T_{\mathrm{m}}}.\]

@@

\hypertarget{s3-sensitivity-analysis-of-parameters-in-ibm}{%
\subsection{S3: Sensitivity analysis of parameters in
IBM}\label{s3-sensitivity-analysis-of-parameters-in-ibm}}

\begin{figure}
\centering
\includegraphics{index_files/figure-latex/unnamed-chunk-6-1.pdf}
\caption{Figure S2.1: Evolution of both male search (blue) and female
choice (red) under different combinations of the mortality rates
\(\mu_{\mathrm{in}}\) and \(\mu_{\mathrm{out}}\) (mortality in time-in
and out, respectively). The y-axis is the threshold fitness that leads
to evolution of male search (blue) or female choice (red). The results
show noise, but no correlation between the value of the mortality
parameters and the propensity for male search and/or female choice to
evolve. For each of the \(25 \times 20\) combinations of
\(\mu_{\mathrm{in}}\) and \(\mu_{\mathrm{out}}\), 3000 replicate
simulations were run.}
\end{figure}

\begin{figure}
\centering
\includegraphics{index_files/figure-latex/unnamed-chunk-7-1.pdf}
\caption{Figure S2.3 (\(T_{\mathrm{f}} = 10.0\)): The coevolution of
male search and female choosiness as a function of nuptial gift search
time (\(\alpha\)). Points show where the lower 95\% confidence interval
of female choosiness (red) and male search (blue) exceeds zero,
indicating evolution of choosiness or nuptial gift search. Each point
includes data from 3000 replicate simulations with identical starting
conditions. Up to 3000 interactions occur between individuals in each
time step (\(\psi = 3\)), potentially resulting in a mating interaction.
The number of individuals in the population remained at or near carrying
capacity of \(K = 1000\). Expected female processing time was set to
\(T_{\mathrm{f}}=10.0\) time steps, and \(\gamma\) and \(\alpha\) values
in the range {[}0.1, 2.0{]} and {[}0.0, 2.0{]}, respectively, were
used.}
\end{figure}

>>>>

We develop a general framework for investigating the evolution of
nuptial gift-giving and choosiness using a time-in, time-out modelling
approach and an individual-based model
(\protect\hyperlink{ref-Clutton-Brock1992}{Clutton-Brock \& Parker,
1992}). Specifically, we derive conditions under which selection will
favour male search for nuptial gifts and female rejection of gift-less
males. We show that selection for searching and choosiness depends on
whether a threshold fitness value of the nuptial gift is exceeded. Our
model demonstrates the importance of nuptial gift cost, sex ratio, and
mate encounter rate in determining the threshold above which selection
will favour the evolution of nuptial gift-giving. Importantly, we show
that the threshold value differs for male searching and female
choosiness. We complement the predictions of our analytical model by
formulating an individual-based model, which further supports the main
theoretical results. We apply our model to an example system with
nuptial gifts, the nursery web spider \emph{Pisaura mirabilis}, where we
use experimental data to estimate a key model parameter. Our results
provide a general framework for understanding why nuptial gift-giving
evolves in some systems and not in others, how the evolution of nuptial
gift-giving can give rise to sexual conflict, and it provides insight
into the evolution of worthless and deceitful nuptial gifts, which occur
in several different taxonomic groups
(\protect\hyperlink{ref-LeBas2005}{LeBas \& Hockham, 2005};
\protect\hyperlink{ref-Ghislandi2014}{Ghislandi \emph{et al.}, 2014}).

@@

\begin{figure}
\includegraphics[width=1\linewidth]{inst/img/conceptual_figure} \caption{Conceptual figure inspired by Kokko and Ots (2006) illustrating how the modelling framework maps onto an example of a system wherein nuptial gifts are used, here \textit{Pisaura mirabilis}. Males have a probability of obtaining a nuptial gift while in time-out, which will affect their probability of mating while in time-in. They return to the mating pool (time-in) at a rate determined by the time spent searching for a nuptial gifts ($T_{\mathrm{m}}$) and leave the mating pool (i.e. enter time-out) following the female encounter rate, which is dependent on the ratio of males to females ($\beta$) and the encounter rate ($R$). The choosy sex (females) enter the mating pool at a rate depending on the time spent processing offspring ($T_{\mathrm{f}}$) and leave the mating pool (i.e. enter time-out) at a rate that is dependent on $\beta$ and $R$. Males and females undergo mortality $\mu$ during time-in and time-out. Image left to right: (1) male \textit{P. mirabilis}. (2) male \textit{P. mirabilis} presenting nuptial gift (white) to female. (3) Female \textit{P. mirabilis} protecting offspring. Photos: Alamy.}\label{fig:unnamed-chunk-2}
\end{figure}

@@

The probability that a male who searches for an interval
\(T_{\mathrm{m}}\), and does not die during this period, finds a nuptial
gift is given by,

\[G = 1 - e^{-\frac{1}{\alpha}T_{m}}.
\tag{1}
\]

In Eq. 1, \(\alpha\) defines expected search time before encountering a
nuptial gift. Thus, the probability of finding a nuptial gift is higher
the more time is spent searching. During time-in, individuals encounter
conspecifics at a rate of \(R\). A focal individual will therefore
encounter conspecifics of the opposite sex at a rate of \(R/2\) if the
ratio of males to females in the mating pool (\(\beta\)) is equal. More
generally, males will be encountered at a rate of \(R\beta/(\beta+1)\),
and females will be encountered at a rate of \(R/(\beta+1)\). An example
of how the structure of the time-in time-out model applies to a system
with nuptial gift-giving is given in Figure 1. We assume that mating
with a nuptial gift increases expected offspring production by an
increment of \(\gamma\).

Mortality of \(\mu_{\mathrm{f}}\) and \(\mu_{\mathrm{m}}\) for females
and males, respectively, occurs both in and out of the mating pool.
Following Kokko \& Ots (\protect\hyperlink{ref-Kokko2006}{2006}), we
assume \(\mu_{\mathrm{f}} = \mu_{\mathrm{m}} = 1\). Assuming identical
mortality both in and out of the mating pool means that individual
phenotypes (search strategy and choosiness) do not affect mortality.
This allows us to use a rate-maximisation model wherein maximisation of
reproductive rate while alive is equivalent to maximising lifetime
reproductive success (\protect\hyperlink{ref-Kokko2008}{Kokko \&
Jennions, 2008}; \protect\hyperlink{ref-Jennions2017}{Jennions \&
Fromhage, 2017}). Here maximisation of lifetime reproductive success is
the same as maximising fitness because we do not model social
interaction between relatives (\protect\hyperlink{ref-Mylius1995}{Mylius
\& Diekmann, 1995}; \protect\hyperlink{ref-Fromhage2024}{Fromhage
\emph{et al.}, 2024}). It should therefore be noted that if mortality
differs inside and outside of the mating pool, reproductive rate while
alive will no longer be equivalent to fitness. Nevertheless, we also
investigated the evolution of both traits when mortality rates in and
out of the mating pool differed substantially (see Supporting
Information S1).

@@

During time-out, males can search for a nuptial gift. Males can adopt
one of two strategies; either search or do not search for a nuptial
gift. Males with the searching strategy continue to search until they
find a nuptial gift, while males that do not search immediately re-enter
the mating pool. Consequently, time searching for a nuptial gift will
come at the cost of mating opportunities but might increase fecundity.
We therefore need to model the expected length of time spent outside of
the mating pool for males that search for nuptial gifts
(\(E[T_{\mathrm{m}}]\)), which equals \(\alpha\). We can integrate
search time \(t\) over the rate at which nuptial gifts are encountered
(\(\exp(-1/\alpha)\)) to show \(E[T_{\mathrm{m}}] = \alpha\),

@@

When this inequality holds, males are favoured to search until they find
a nuptial gift, which would result in an average search time of
\(\alpha\). If the male trait is instead continuous (i.e., males search
for time period \(T_{\mathrm{m}}\)), it can be shown that the same
threshold can be reached by evaluating the partial derivative of the
male fitness function (Supporting Information S2). Hence, the thresholds
are consistent under different assumptions concerning male searching
strategy. Selection will cause males to search for nuptial gifts if the
fitness increase to offspring exceeds the product of search time and
female encounter rate.

@@

We then evaluate the conditions under which
\(W_{\mathrm{F_{G}}} > W_{\mathrm{F_{G,L}}}\). We can isolate \(\gamma\)
to determine how much fecundity must be increased to make choosiness
beneficial (\(\gamma_{\mathrm{f}}\)),

@@

In Eq. 6, \(P_{IM}(t)\) and \(P_{IF}(t)\) are the probabilities of
finding a male and female in time-in at time \(t\), respectively. There
is no closed form solution to the operational sex ratio, so we used
recursion to calculate \(\beta\) values for a given \(T_{\mathrm{f}}\),
\(T_{\mathrm{m}}\), and \(R\) (see Supporting Information S3),

@@

\[G^{i} = 1 - e^{-\frac{1}{\alpha}T^{i}_{\mathrm{m}}}.\]

@@

We ran simulations in which male search time and female choosiness
evolved from an ancestral state of no searching and no choosiness. In
all simulations, \(N\) was initialised at 1000 and \(K = 1000\).
Simulations ran for \(t_{max} = 10000\) time steps. We set
\(T_{\mathrm{f}} = 2\), \(\psi = 3\), and \(\lambda = 1\) for all
simulations, and we simulated across a range of
\(\alpha = \{0.1, 0.2, ..., 1.9, 2.0\}\) and
\(\gamma = \{0, 0.1, ..., 1.9, 2.0\}\) parameter values for 5000
replicates. In a sensitivity analysis, we also re-ran simulations for
values of \(T_{\mathrm{f}} = 10\) and \(\psi = \{1, 2, 4, 6 \}\) (see
Supporting Information S4), and simulated the evolution of male nuptial
gift searching and female choosiness (fixing
\(T_{\mathrm{m}} = \alpha\)) separately (see Supporting Information S4).

We additionally ran simulations in which males adopted one of two search
strategies: either search in time-out until a gift is obtained, or do
not search at all. This simulations were therefore more closely aligned
with the assumptions of our analytical model, but results did not differ
qualitatively (Supporting Information S5). The strategy adopted by a
male in these simulations was based on an underlying trait for the
probability of nuptial gift search. Like \(\phi^{i}\), this trait was
underpinned by a diploid genotype in which alleles had additive effects.

Summary statistics for mean trait values, population size, sex ratios,
proportion of females and males in and out of the mating pool, and mean
number of encounters per female and male within the mating pool were all
calculated in the last time step. The C code used for simulating these
IBMs also allows for the reporting of summary statistics in each time
step. Additionally, it can simulate explicit space and individual
movement through the landscape. A neutral evolving trait was also
modelled to ensure that the code functioned as intended, and processes
were compartmentalised into individual functions to facilitate code
testing. All code is publicly available on GitHub
(\url{https://github.com/bradduthie/nuptial_gift_evolution}).

@@

\begin{figure}
\centering
\includegraphics{index_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{Fecundity increment above which males increase their fitness by
searching for nuptial gifts (blue lines; Eq. 2) and females increase
their fitness by rejecting males that do not offer gifts (red lines; Eq.
3). Parameter space includes areas in which males do not search for
nuptial gifts and females are not choosy (A), males search but females
are not choosy (B), females would be choosy but males do not search (C),
and males search and females are choosy (D). Arrows in panel a indicate
the effect of increasing encounter rate (\(R\)), female time-out
(\(T_{\mathrm{f}}\)), and male search time (\(T_{\mathrm{m}}\)). As an
example, curves for \(T_{\mathrm{f}} = 2\), and
\(T_{\mathrm{m}} = \alpha\) are shown for values of \(R=2\) (panel a)
and \(R = 1\) (panel b). Females are assumed to be the choosy sex, which
is maintained as long as \(\alpha < T_{\mathrm{f}}\).}
\end{figure}

For males, if nuptial gifts are not abundant and thus require a long
time to find (i.e., high \(\alpha\)), or if males encounter many females
per unit time (i.e., high \(R / (1+\beta)\)), then the nuptial gift must
result in a high fitness increment for selection to favour searching.
For females, if \(\gamma\) is sufficiently high, fitness is increased by
rejecting males without gifts and mating only with males that provide
nuptial gifts. As offspring processing time (\(T_{\mathrm{f}}\)), mate
encounter rate (\(R\beta / (\beta + 1)\)), or the probability of a male
finding a nuptial gift (\(1 - \exp(-T_{\mathrm{m}}/\alpha)\) ) decrease,
the fecundity increment above which selection will favour choosiness
(\(\gamma_{\mathrm{f}}\)) increases. This can be understood intuitively
by realising that rejecting a prospective male represents an opportunity
cost for the female.

@@

We show that the threshold nuptial gift value at which females are
favoured to express choosiness for nuptial gifts is rarely equivalent to
the threshold value at which males are favoured to search for nuptial
gifts, potentially leading to sexual conflict
(\protect\hyperlink{ref-Arnqvist2005a}{Arnqvist \& Rowe, 2005};
\protect\hyperlink{ref-Oliveira2008}{Oliveira \emph{et al.}, 2008}).
Here, we are defining sexual conflict as occurring when interactions
between sexes result in situation where both sexes cannot achieve an
optimal outcome simultaneously
(\protect\hyperlink{ref-Parker2006}{Parker, 2006}). As an example,
sizable areas of parameter space exists wherein the female optimum would
be to exhibit preference for (and receive) nuptial gifts, while the male
optimum is to not search for (and give) nuptial gifts (see Figure 2a,
Zone C). This will lead to encounters where gift-less males would
benefit the most from mating, but females would benefit most from mating
and receiving a gift (i.e., evolutionary interests do not always overlap
with respect to mating strategy between the sexes). In many systems,
ecological variables such as search time required to find a nuptial gift
will likely depend on prey abundance, which can vary substantially with
time in some species with nuptial gift-giving
(\protect\hyperlink{ref-Ghislandi2018}{Ghislandi \emph{et al.}, 2018}).
Since several ecological variables likely affect the value of these
thresholds, our results can be seen as providing some formalised
description of why nuptial gift-giving only occurs in some but not all
systems.

At first, the analytical model seems to suggests that nuptial gifts must
cause a very high fitness increase (approximately 25\%) before male
search and female choosiness is favoured by selection (Figure 2).
Similarly the IBM model seems to suggest that a fitness benefit of
approximately 50\% is required (see Figure 3). However, it is important
to recognise that these thresholds depend on multiple parameters. For
example, if female processing time (\(T_{\mathrm{f}}\)) is high, the
female threshold for choosiness with respect to \(\gamma\) drops such
that male search and female choosiness are favoured at lower \(\gamma\).
If \(T_{\mathrm{f}}\) is sufficiently high, then an initially rare
gift-giving trait might be favoured by selection even if the fitness
benefit of a nuptial gift is low (assuming variation in choosiness for
gifts existed among females). The effect that nuptial gifts have on
fitness might vary across species, or even populations. Effects on
female fecundity have been estimated in crickets, fireflies,
butterflies, and spiders, but these estimates vary considerably. For
example, one study of crickets (\emph{Oecanthus nigricornis}) found that
more nuptial feeding could increase female reproductive lifespan by
approximately 50\% (\protect\hyperlink{ref-Brown1997}{Brown, 1997}), and
a study on fireflies (\emph{Photinus ignites} and \emph{Ellychnia
corrusca}) reported an increase in fecundity of 73\% for females
consuming multiple nuptial gifts
(\protect\hyperlink{ref-Rooney2002}{Rooney \& Lewis, 2002}). However,
other studies on different systems have reported very varying effects or
even no effect at all (\protect\hyperlink{ref-Bergstrom2002}{Bergström
\& Wiklund, 2002}; \protect\hyperlink{ref-Rooney2002}{Rooney \& Lewis,
2002}; \protect\hyperlink{ref-Maxwell2018}{Maxwell \& Prokop, 2018};
\protect\hyperlink{ref-Gao2019}{Gao \emph{et al.}, 2019}).

@@

\leavevmode\vadjust pre{\hypertarget{ref-Brown1997}{}}%
Brown, W.D. (1997) Courtship feeding in tree crickets increases
insemination and female reproductive life span. \emph{Animal Behaviour},
\textbf{54}, 1369--1382.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Fromhage2024}{}}%
Fromhage, L., Jennions, M.D., Myllymaa, L. \& Henshaw, J.M. (2024)
Fitness as the organismal performance measure guiding adaptive
evolution. \emph{Evolution}, \textbf{78}, 1039--1053.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Jennions2017}{}}%
Jennions, M.D. \& Fromhage, L. (2017) Not all sex ratios are equal: The
fisher condition, parental care and sexual selection.
\emph{Philosophical Transactions of the Royal Society B: Biological
Sciences}, \textbf{372}, 20160312.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Kokko2008}{}}%
Kokko, H. \& Jennions, M.D. (2008) Parental investment, sexual selection
and sex ratios. \emph{Journal of Evolutionary Biology}, \textbf{21},
919--948.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Mylius1995}{}}%
Mylius, S.D. \& Diekmann, O. (1995) On evolutionarily stable life
histories, optimization and the need to be specific about density
dependence. \emph{Oikos}, 218--224.

@@

\hypertarget{s1-sensitivity-analysis-of-parameters-in-ibm}{%
\subsection{S1: Sensitivity analysis of parameters in
IBM}\label{s1-sensitivity-analysis-of-parameters-in-ibm}}

\begin{figure}
\centering
\includegraphics{index_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Figure S2.1: Evolution of both male search (blue) and female
choice (red) under different combinations of the mortality rates
\(\mu_{\mathrm{in}}\) and \(\mu_{\mathrm{out}}\) (mortality in time-in
and out, respectively). The y-axis is the threshold fitness that leads
to evolution of male search (blue) or female choice (red). The results
show noise, but no correlation between the value of the mortality
parameters and the propensity for male search and/or female choice to
evolve. For each of the \(25 \times 20\) combinations of
\(\mu_{\mathrm{in}}\) and \(\mu_{\mathrm{out}}\), 3000 replicate
simulations were run.}
\end{figure}

\begin{figure}
\centering
\includegraphics{index_files/figure-latex/unnamed-chunk-6-1.pdf}
\caption{Figure S2.3 (\(T_{\mathrm{f}} = 10.0\)): The coevolution of
male search and female choosiness as a function of nuptial gift search
time (\(\alpha\)). Points show where the lower 95\% confidence interval
of female choosiness (red) and male search (blue) exceeds zero,
indicating evolution of choosiness or nuptial gift search. Each point
includes data from 3000 replicate simulations with identical starting
conditions. Up to 3000 interactions occur between individuals in each
time step (\(\psi = 3\)), potentially resulting in a mating interaction.
The number of individuals in the population remained at or near carrying
capacity of \(K = 1000\). Expected female processing time was set to
\(T_{\mathrm{f}}=10.0\) time steps, and \(\gamma\) and \(\alpha\) values
in the range {[}0.1, 2.0{]} and {[}0.0, 2.0{]}, respectively, were
used.}
\end{figure}

\begin{figure}
\centering
\includegraphics{index_files/figure-latex/unnamed-chunk-7-1.pdf}
\caption{Figure S2.4 (\(\psi = 1\)): The coevolution of male search and
female choosiness as a function of nuptial gift search time
(\(\alpha\)). Points show where the lower 95\% confidence interval of
where male search (blue) exceeds zero, indicating evolution of
choosiness or nuptial gift search. Each point includes data from 3000
replicate simulations with identical starting conditions. Up to 1000
interactions occur between individuals in each time step (\(\psi = 1\)),
potentially resulting in a mating interaction. The number of individuals
in the population remained at or near carrying capacity of \(K = 1000\).
Expected female processing time was set to \(T_{\mathrm{f}}=2\) time
steps, and \(\gamma\) and \(\alpha\) values in the range {[}0.0, 2.0{]}
and {[}0.1, 2.0{]}, respectively, were used.}
\end{figure}

\begin{figure}
\centering
\includegraphics{index_files/figure-latex/unnamed-chunk-8-1.pdf}
\caption{Figure S2.5 (\(\psi = 2\)): The coevolution of male search and
female choosiness as a function of nuptial gift search time
(\(\alpha\)). Points show where the lower 95\% confidence interval of
where male search (blue) exceeds zero, indicating evolution of
choosiness or nuptial gift search. Each point includes data from 3000
replicate simulations with identical starting conditions. Up to 2000
interactions occur between individuals in each time step (\(\psi = 2\)),
potentially resulting in a mating interaction. The number of individuals
in the population remained at or near carrying capacity of \(K = 1000\).
Expected female processing time was set to \(T_{\mathrm{f}}=2\) time
steps, and \(\gamma\) and \(\alpha\) values in the range {[}0.0, 2.0{]}
and {[}0.1, 2.0{]}, respectively, were used.}
\end{figure}

\begin{figure}
\centering
\includegraphics{index_files/figure-latex/unnamed-chunk-9-1.pdf}
\caption{Figure S2.6 (\(\psi = 4\)): The coevolution of male search and
female choosiness as a function of nuptial gift search time
(\(\alpha\)). Points show where the lower 95\% confidence interval of
where male search (blue) exceeds zero, indicating evolution of
choosiness or nuptial gift search. Each point includes data from 3000
replicate simulations with identical starting conditions. Up to 4000
interactions occur between individuals in each time step, potentially
resulting in a mating interaction (\(\psi = 4\)). The number of
individuals in the population remained at or near carrying capacity of
\(K = 1000\). Expected female processing time was set to
\(T_{\mathrm{f}}=2\) time steps, and \(\gamma\) and \(\alpha\) values in
the range {[}0.0, 2.0{]} and {[}0.1, 2.0{]}, respectively, were used.}
\end{figure}

\begin{figure}
\centering
\includegraphics{index_files/figure-latex/unnamed-chunk-10-1.pdf}
\caption{Figure S2.7 (\(\psi = 6\)): The coevolution of male search and
female choosiness as a function of nuptial gift search time
(\(\alpha\)). Points show where the lower 95\% confidence interval of
where male search (blue) exceeds zero, indicating evolution of
choosiness or nuptial gift search. Each point includes data from 3000
replicate simulations with identical starting conditions. Up to 6000
interactions occur between individuals in each time step, potentially
resulting in a mating interaction (\(\psi = 6\)). The number of
individuals in the population remained at or near carrying capacity of
\(K = 1000\). Expected female processing time was set to
\(T_{\mathrm{f}}=2\) time steps, and \(\gamma\) and \(\alpha\) values in
the range {[}0.0, 2.0{]} and {[}0.1, 2.0{]}, respectively, were used.}
\end{figure}

\hypertarget{s2-alternative-derivation-of-male-fitness-threshold}{%
\subsection{S2: Alternative derivation of male fitness
threshold}\label{s2-alternative-derivation-of-male-fitness-threshold}}

In the main text, we assumed that males made the decision to search or
not search for a nuptial gift. The expected length of time for which
searching males are expected to remain outside of the mating pool is
\(E[T_{\mathrm{m}}] = \alpha\) (see Model). Alternatively, we can assume
that males search for a pre-determined period of \(T_{\mathrm{m}}\) and
spend this full duration of \(T_{\mathrm{m}}\) in the time-out phase,
even if they succeed in finding a nuptial gift (as we assumed in our
IBM). The probability that a male obtains a nuptial gift during this
time is modelled in Eq. 1,

\[G = 1 - e^{-\frac{1}{\alpha}T_{\mathrm{m}}}.\]

@@
